Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-01-11 Manuscript_Nater_etal.qmd
Added abstract, text to discussion, and some bibs.

TBA

@@

Building on the work of @nilsen2023, we applied a novel integrated population model to data collected through a national-scale participatory monitoring programme to estimate spatial and temporal variation in demography of a culturally important game bird species, the willow ptarmigan. While our study was exploratory in nature, it recovered patterns consistent with ecological and life-history theory including trade-offs between survival and recruitment, and a tendency towards slower life histories at higher latitudes and altitudes. Space-for-time substitution also provided the statistical power necessary for the analysis to provide evidence for the alternative prey hypothesis, i.e. ptarmigans benefiting from high abundance of alternative prey for their predators. <!-- To me, this little paragraph has "abstract vibes", or maybe "conclusion vibes". I'll leave it here for now but note that it may get moved eventually.

## Abundance and vital rates across space and time

@@

The wide spatial distribution of the line transect monitoring afforded us the opportunity to explore variation in population density and vital rates across a relatively large spatial extent. Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA]. These areas are characterized by more extreme climatic conditions, boasting cold temperatures and – in the North – a short growing season. Resulting reduced primary production limits food availability and as ptarmigan are income breeders using food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected. <!-- @ERLEND do we have some more thoughts to add here? -->

We found increasing population trends over recent years in over half of the reporting districts [@fig-lamA]. <!-- relate this to changes in hunting management in recent years? Any other factors we think could have contributed? --> Nonetheless, recent population declines were also evident in some areas, and particularly in the mountains in central Norway [@fig-lamA]. <!-- What could be going on here? It's also interesting that while some of these areas have declines over the most recent years, that usually follows on a period of increase, or at least stability, up to around 2016 -->

While we found the highest population growth rates in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A], our analysis did not provide evidence for a strong relationship between density and growth rates across space. However, this is unsurprising as we did find substantial spatial differences in demography, which suggests that carrying capacity too varies over space. Overall, annual recruitment rates showed a spatial pattern that was inversely related to that of annual survival, with lower recruitment rates co-occurring with higher survival rates (and lower densities) and vice-versa [@fig-paramCorrs B], as we would expecte based on life history theory [@stearns1992]. Furthermore, the areas with relatively lower recruitment and higher survival were primarily found in the northern parts of Norway and in the mountainous region in the central part of the country, which aligns with previous studies finding relatively slower bird life histories in alpine / high altitude areas [e.g. @sandercock2005; @bears2009; @Wilson_Martin_2011; @alice2016]. <!-- What about the latitude? Do we know of any studies looking at latitudinal variation in life histories? When I last searched, I only found studies focusing on altitude -->

<!-- Temporal perspective -->
Population densities in all areas were subject to substantial variation across years, featuring periods of stability, increase, and decrease (SF "TimeSeries_PopDens1.pdf"). In most areas, there were also strong year-by-year fluctuations in population density on top of longer-term trends. Some of the resulting "high density years" were highly synchronized across large spatial scales, such as the years 2011, 2013, and 2018. Taking a closer look, we find that these are years that are characterized by high recruitment (SF "TimeSeries_rRep.pdf") and low subsequent survival (SF "TimeSeries_pSurv.pdf"), which often resulted in steep population declines towards the following year. The fact that these same years also match up with observed peaks in rodent abundance in many areas provides evidence for the Alternative Prey Hypothesis (APH). The AHP stipulates that high abundance of alternative prey (rodents, in this case) for common predators leads to population growth, and is well-supported throughout the literature for a range of taxa [e.g., @hagen1952; @Kjellander_Nordström_2003; @Reif_Tornberg_Jungell_Korpimäki_2001], including willow ptarmigan [@bowler2020]. <!-- Maybe another sentence or two about the low survival following the peak, which could be either due to negative density-dependence leading to resource competition among ptarmigan, or due to the rodent peak allowing high recruitment also for predators, which then in turn will have increased impacts on all prey populations in the subsequent year? -->

Taken together, this alludes to the large role small rodents play in shaping ptarmigan demography and population dynamics and is further supported by the largely positive effects of rodent abundance on recruitment estimated by our model (@fig-paramCorrs C). Once more, the highest latitude and highest altitude areas stood out, sporting the strongest rodent effects (SF "betaR_Map.pdf"). This may be linked to the overall low ptarmigan densities in these areas, as e.g. @Nielsen_1999 found evidence that gyrfalcon predation on ptarmigan was most impactful when ptarmigan densities were low. 
<!-- This previous sentence is adapted from a suggestion by James, but I feel like we need some more explanation/arguments here to clarify this. Perhaps also some other and/or more specific hypotheses= -->
Notably, the conclusive estimation of overall positive effects of rodent abundance on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas. When @nilsen2023 fit the IDSM to data from only a single area, they were unable to obtain a reliable estimate for the rodent effect due to limited statistical power. Consequently, the space-for-time substitution that comes with extending the model across mulitple area allows estimation of covariate effects that otherwise cannot be estimated, and opens up for future possibilities for studying effects of not just rodents, but also other environmental drivers on ptarmigan population dynamics. 
For example, [...] <!-- @ERLEND here we could add some thoughts on other covariates that may matter and are realistic/relevant/realistic to include in future studies. Best if it kind of naturally leads over to (and does not overlap too much with) the next paragraph  -->

<!-- Variation beyond rodents -->
Extending the present analysis by including additional covariates may also help with better understanding the mechanisms underlying the large portion of demographic rate variation that could only be attributed to random variation so far. This is the case especially for the relatively large residual variation [@fig-varDecomp] but also relevant for constant spatial and shared temporal variation. Interestingly, we found that a larger portion of variation in recruitment was allocated to spatial than to temporal factors, while the opposite was the case for survival. [...]
<!-- Any hypotheses/thoughts related to large- vs. local-scale drivers to explain this? -->

@@

Management decisions made at the resolution of large geo-political boundaries (e.g., Norway) run a high risk of being inadequate when there is substantial spatial variation in demographic processes and population dynamics, as is the case for willow ptarmigan. In Norway, willow ptarmigan -- and small game in general -- is managed [...] <!-- Brief summary of how and at what spatial scales patrmigan/small game management works in Norway -->
The results from ur study suggests that management decisions should be spatially refined and ideally informed by knowledge about the "local" population processes. [...] <!-- Some specific implications of the results for management of hunting and of habitat (e.g. better managed hunting is good because we do see increases in majority of local populations, but the mountain areas have had recent declines that may call for additional local measures, these measures may need to be adjusted since the mountain populations appear to have slower life histories and likely different sensitivities, etc.) -->

Besides pointers on how to adjust management of ptarmigan and their habitat, our results also provided some insights into the value, and possibly opportunities for improving the monitoring programme. First and foremost, our study demonstrates the tremendous potential lying in coordinating structured monitoring using a common protocol and training programme, as well as data processing pipeline. These were indeed the prerequisites that allowed us to easily and efficiently integrate data collected across the entire country in a joint analysis, and draw inference on spatio-temporal variation in demography and population dynamics at a large scale but also with good spatial resolution. 
While overall less variable across space and time than vital rates, differences in detection probabilities were nonetheless evident (SFs "Avg_detect_Map.pdf" and "TimeSeries_pDetect.pdf") and may help with mapping out potential for improvement in the monitoring programme. Particularly, we found generally lower detection probabilities in the northern half of Norway. [...] <!-- Some thoughts on why that might be. Could be related to landscape features and/or available observer pool. Any suggestions on what could be done to improve detection in the North, or at least "make up" for it? -->
Ultimately, increased detection probability would contribute to obtaining more precise estimates of both population density and demographic rates, which -- in turn -- would be of great value in particular in areas with relatively low population densities, low number of transects, and less years of data.

@@

-   Outlook: integration with / coupling to models for harvest to develop tools for (automatic) calculation of e.g. hunting quotas <!-- @JAMES: This one is for you. Some ruminations on how and why it would be great to formally model harvest, integration with hunting data, maybe eventually develop a tool to calculate quotas (I know this has been developed for other game species), etc. -->

@@

TBA ... Additionally, we would like to thank Bernardo Brandão Niebuhr dos Santos for his help with spatial data visualization. <!-- @ERLEND: you maybe know best all the people/institutions we need to thank here --> <!-- @JAMES: should we be acknowledging your fullbright scholarship which made the collab possible? If so, please add details. -->

>>>>

Quantifying the temporal and spatial variation in animal demography is a central theme of ecology and important for directing policy. However, field sampling at large spatial extents for long periods of time is cost prohibitive and often politically untennable. Citizen science programs can alleviate these issues by harnessing stakeholders to increase the spatial and temporal resolution of datasets. Yet, many citizen science programs lack data on individual animals making it difficult to model demography. Integrated population models (IPM) offer promise to use multiple streams of data on populations or multiple aspects of a single dataset to estimate key vital rates for populations or communities. We leveraged a large-scale citizen science monitoring program---the "Hønsefuglportalen"--- to estimate spatial and temporal variation in demography for willow ptarmigan (*Lagopus lagopus*). The core field sampling approach was line-transect distance sampling with age-specific detections---the latter critical to modelling the population dynamics in the IPM. We explored latitudinal and longitudinal variation in recruitment, survival, and density; patterns of density dependence; and the relationship between recruitment and rodent occupancy using an IPM in a Bayesian framework following open science principles. Overall results were consistent with expectations of ecological theory. Slower life history patterns were observed at higher latitudes and altitudes portending differential effects of climate change on ptarmigan across their range. Observed patterns between rodents and recruitment were consistent with the alternative prey hypothesis (APH); however, spatial variation in the directionality and magnitude of effect were substantial. The robustness of the hierarchical structure of our model was instrumental in elucidating these patterns. Our study highlights the power of citizen science, IPMs, and open science principles to estimate spatial and temporal variation in the demography for a culturally important species.

@@

Building on the work of @nilsen2023, we applied a novel integrated population model to data collected through a national-scale participatory monitoring programme to estimate spatial and temporal variation in demography of a culturally important game bird species, the willow ptarmigan. While our study was exploratory in nature, it recovered patterns consistent with ecological and life-history theory including trade-offs between survival and recruitment, and a tendency towards slower life histories at higher latitudes and altitudes. Space-for-time substitution also provided the statistical power necessary for the analysis to provide evidence for the alternative prey hypothesis, i.e. ptarmigans benefiting from high abundance of alternative prey for their predators. <!-- To me, this little paragraph has "abstract vibes", or maybe "conclusion vibes". I'll leave it here for now but note that it may get moved eventually. 
<!-- Matter of style but this is typically how I do it but your call. :-) >
## Abundance and vital rates across space and time

@@

The wide spatial distribution of the line transect monitoring afforded us the opportunity to explore variation in population density and vital rates across a relatively large spatial extent. Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA]. These areas are characterized by more extreme climatic conditions, boasting cold temperatures and -- in the North -- a short growing season. Resulting reduced primary production limits food availability and as ptarmigan are income breeders using food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected. <!-- @ERLEND do we have some more thoughts to add here? -->We found increasing population trends over recent years in over half of the reporting districts [@fig-lamA]. <!-- relate this to changes in hunting management in recent years? Any other factors we think could have contributed? --> Nonetheless, recent population declines were also evident in some areas, and particularly in the mountains in central Norway [@fig-lamA]. <!-- What could be going on here? It's also interesting that while some of these areas have declines over the most recent years, that usually follows on a period of increase, or at least stability, up to around 2016 -->

<!-- Erlend will probably have better ideas than I.-->

Density dependence is an important process to understand in ecology but more so for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @Aanes_Engen_SÆther_Willebrand_Marcström_2002; @willebrand2001]. Negative density dependence has been found in numerous gallinaceous birds such as northern bobwhites [*Colinus virginianus*; [@mcconnell2018]\], \[*Perdix perdix*; [@bro2003]\], and wild turkeys \[*Meleagris gallopavo*; [@mcghee2007]]*.* <!--# Got the tasty ones in here for you -->While we found the highest population growth rates in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A], our analysis did not provide evidence for a strong relationship between density and growth rates across space. However, this is unsurprising as we did find substantial spatial differences in demography, which suggests that carrying capacity too varies over space. Also, support for density dependence in willow ptarmigan has been mixed. @myrberget1988 observed no change in productivity despite a 50% decrease in abundance. Conversely, @pedersen2004 found strong overwinter negative density dependence and they posited that immigration was the vital rate that responded to changes in density. Low survival rates appeared to have a one-year lag following abundance peaks. This pattern could be a result of negative density dependence that went undetected when looking at the ensemble of samples across the 40 areas. Negative density dependence has been shown in willow ptarmigan populations which may explain explain reduced survival following years with elevated densities [@pedersen2004].

<!--# Not happy with the DD para yet but I can work on Jan 11 -->

<!-- Maybe another sentence or two about the low survival following the peak, which could be either due to negative density-dependence leading to resource competition among ptarmigan, or due to the rodent peak allowing high recruitment also for predators, which then in turn will have increased impacts on all prey populations in the subsequent year? -->

Overall, annual recruitment rates showed a spatial pattern that was inversely related to that of annual survival, with lower recruitment rates co-occurring with higher survival rates (and lower densities) and vice-versa [@fig-paramCorrs B], as we would expect based on life history theory [@stearns1992]. Furthermore, the areas with relatively lower recruitment and higher survival were primarily found in the northern parts of Norway and in the mountainous region in the central part of the country, which aligns with previous studies finding relatively slower bird life histories in alpine / high altitude areas [e.g. @sandercock2005; @bears2009; @Wilson_Martin_2011; @alice2016]. <!-- What about the latitude? Do we know of any studies looking at latitudinal variation in life histories? When I last searched, I only found studies focusing on altitude -->

<!-- Temporal perspective -->

Population densities in all areas were subject to substantial variation across years, featuring periods of stability, increase, and decrease (SF "TimeSeries_PopDens1.pdf"). In most areas, there were also strong year-by-year fluctuations in population density on top of longer-term trends. Some of the resulting "high density years" were highly synchronized across large spatial scales, such as the years 2011, 2013, and 2018. Taking a closer look, we find that these are years that are characterized by high recruitment (SF "TimeSeries_rRep.pdf") and low subsequent survival (SF "TimeSeries_pSurv.pdf"), which often resulted in steep population declines towards the following year. The fact that these same years also match up with observed peaks in rodent abundance in many areas provides evidence for the Alternative Prey Hypothesis (APH). The AHP stipulates that high abundance of alternative prey (rodents, in this case) for common predators leads to population growth, and is well-supported throughout the literature for a range of taxa [e.g., @hagen1952; @Kjellander_Nordström_2003; @Reif_Tornberg_Jungell_Korpimäki_2001], including willow ptarmigan [@bowler2020]. Conversely, @Nyström_etal_2006 suggested that gyrfalcons (a specialized ptarmigan predator) did not respond to rodent populations and did not switch to alternative prey when ptarmigan populations are low. Nonetheless, other predators such as fox could prey switch in response to changing availability of rodents and ptarmigan as shown by @Breisjøberget_etal_2018 and @bowler2020. However, statistical associations in our study and others is not causal and alternative explanations for this pattern are plausible. The positive association between rodents and ptarmigan could be caused by bottom-up processes such as food availability unrelated to predation. Future work should develop causal models and manipulative experiments to establish causal relationships among top-down and bottom-up processes.

Taken together, this alludes to the large role small rodents play in shaping ptarmigan demography and population dynamics and is further supported by the largely positive effects of rodent abundance on recruitment estimated by our model (@fig-paramCorrs C). Once more, the highest latitude and highest altitude areas stood out, sporting the strongest rodent effects (SF "betaR_Map.pdf"). This may be linked to the overall low ptarmigan densities in these areas, as e.g. @Nielsen_1999 found evidence that gyrfalcon predation on ptarmigan was most impactful when ptarmigan densities were low. <!-- This previous sentence is adapted from a suggestion by James, but I feel like we need some more explanation/arguments here to clarify this. Perhaps also some other and/or more specific hypotheses= --> Notably, the conclusive estimation of overall positive effects of rodent abundance on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas. When @nilsen2023 fit the IDSM to data from only a single area, they were unable to obtain a reliable estimate for the rodent effect due to limited statistical power. Consequently, the space-for-time substitution that comes with extending the model across multiple area allows estimation of covariate effects that otherwise cannot be estimated, and opens up for future possibilities for studying effects of not just rodents, but also other environmental drivers on ptarmigan population dynamics. For example, \[...\] <!-- @ERLEND here we could add some thoughts on other covariates that may matter and are realistic/relevant/realistic to include in future studies. Best if it kind of naturally leads over to (and does not overlap too much with) the next paragraph  -->

<!-- Variation beyond rodents -->

Extending the present analysis by including additional covariates may also help with better understanding the mechanisms underlying the large portion of demographic rate variation that could only be attributed to random variation so far. This is the case especially for the relatively large residual variation [@fig-varDecomp] but also relevant for constant spatial and shared temporal variation. Interestingly, we found that a larger portion of variation in recruitment was allocated to spatial than to temporal factors, while the opposite was the case for survival. \[...\] <!-- Any hypotheses/thoughts related to large- vs. local-scale drivers to explain this? -->

<!-- This is interesting I will do some reading on this and add some text on Jan 11-->

@@

Management decisions made at the resolution of large geo-political boundaries (e.g., Norway) run a high risk of being inadequate when there is substantial spatial variation in demographic processes and population dynamics, as is the case for willow ptarmigan. In Norway, willow ptarmigan -- and small game in general -- is managed \[...\] <!-- Brief summary of how and at what spatial scales patrmigan/small game management works in Norway --> The results from ur study suggests that management decisions should be spatially refined and ideally informed by knowledge about the "local" population processes. \[...\] <!-- Some specific implications of the results for management of hunting and of habitat (e.g. better managed hunting is good because we do see increases in majority of local populations, but the mountain areas have had recent declines that may call for additional local measures, these measures may need to be adjusted since the mountain populations appear to have slower life histories and likely different sensitivities, etc.) -->

Besides pointers on how to adjust management of ptarmigan and their habitat, our results also provided some insights into the value, and possibly opportunities for improving the monitoring programme. First and foremost, our study demonstrates the tremendous potential lying in coordinating structured monitoring using a common protocol and training programme, as well as data processing pipeline. These were indeed the prerequisites that allowed us to easily and efficiently integrate data collected across the entire country in a joint analysis, and draw inference on spatio-temporal variation in demography and population dynamics at a large scale but also with good spatial resolution. While overall less variable across space and time than vital rates, differences in detection probabilities were nonetheless evident (SFs "Avg_detect_Map.pdf" and "TimeSeries_pDetect.pdf") and may help with mapping out potential for improvement in the monitoring programme. Particularly, we found generally lower detection probabilities in the northern half of Norway. \[...\] <!-- Some thoughts on why that might be. Could be related to landscape features and/or available observer pool. Any suggestions on what could be done to improve detection in the North, or at least "make up" for it? --> Ultimately, increased detection probability would contribute to obtaining more precise estimates of both population density and demographic rates, which -- in turn -- would be of great value in particular in areas with relatively low population densities, low number of transects, and less years of data.

@@

-   Outlook: integration with / coupling to models for harvest to develop tools for (automatic) calculation of e.g. hunting quotas

    The framework used in our study can be adapted to include the effects of management actions and harvest. Harvest effects on willow ptarmigan has been studied yet uncertainties still exist [@sandercock_etal2011; @Aanes_Engen_SÆther_Willebrand_Marcström_2002; @pedersen2004; @willebrand2001]. Also, there is a dearth of empirical information about the effects of harvest on many exploited species despite decades of continued exploitation.

-   <!-- @JAMES: This one is for you. Some ruminations on how and why it would be great to formally model harvest, integration with hunting data, maybe eventually develop a tool to calculate quotas (I know this has been developed for other game species), etc. -->

-   <!-- Will add more Jan 11 -->

@@

TBA ... We are grateful to the participants in the "Hønsefuglportalen" and their furry friends that collected ptarmigan observations. Additionally, we would like to thank Bernardo Brandão Niebuhr dos Santos for his help with spatial data visualization. J. A. Martin was supported by a Fulbright Fellowship and the Norwegian Institute for Nature Research that facilitated this collaboration. <!-- @ERLEND: you maybe know best all the people/institutions we need to thank here -->
